Empirical Bayes analysis of \RNAseq{} data for detection of gene expression heterosis

### 2015-08-24 response.tex

\comment{Page 6, top.  Is it possible then to assess your proposed method using Laplace versus normal priors to assess if that seems to be contributing to any differences between ShrinkBayes and your proposed method?  Maybe that is the real reason for the small improvements?...., and not the differences in sophistication of the constructed statistical tests?}

\response{We have assessed the use of normal distributions, see Section 3.3.}

\comment{Page 7, line 15.  Why delete the low counts?  Data quality issues?}

\todo{During the course of our research, we were initially estimating hyperparameters based on moment matching of gene-specific parameters estimated using negative binomial regression independently for all genes. This procedure had issues whenever a gene-variety combination had counts that were all zeros, thus leading to an unidentified regression parameter, or too little variability, leading to an unidentified overdispersion parameter. This cut off appeared to fix the problem. None of the current methods suffer from this problem, in particular the \edgeR{} based parameter estimations have a built shrinkage to estimate parameters. We did not change the simulations we were using and thus report the deletion of low counts for reproducible accuracy. We do not believe removing this deletion would change any of our results.}

@@

\comment{Page 8, line 37.  It is not obvious to me why $H_g2*$ (are the two parents the same?) is a necessary component of a heterosis (overdominance) test!?}

\response{It is not necessary, but it is sufficient. If the parents are the same, then the hybrid must have overdominance.}

\comment{Page 9, lines 23-25.  In tandem with another reviewer’s comments, the real question is why?  Is it because of the suboptimality of the test that you contrived for ShrinkBayesA?}

\response{In Section 3.3, we highlighted the differences between ShrinkBayes and the eBayes approaches, but do not try to specifically state which difference is the cause.}

@@

\comment{Page 10, Line 29.  Don’t you mean “nadir” rather than peak?}

\response{No, although we do mean ridge rather than peak, i.e. the count is high for effect size of zero and probabilities below 0.5.}

\comment{Page 11 line 39.  Is it really “independence” or nearly non-identifiability that is the issue?}

\response{We're sorry, but we don't understand this comment. We believe all the parameters in the model are identified and therefore are not sure what non-identifiability is being referenced.}

@@

\response{Fixed.}

Reviewer \#1

\comment{The authors develop a hierarchical negative binomial model for drawing inferences from a composite null hypothesis, that tests gene expression heterosis. Simulations are performed to describe the performance of this new method along with a data analysis demonstrating an application to real data. This manuscript represents a well-conceived and executed methodological development in the statistical analysis of RNA-Seq gene expression data. However, the performance of the new method is underwhelming compared to existing methods, tempering my enthusiasm for the likely impact of this paper.}

\response{We assume that you mean the improvement compared to existing methods is underwhelming. To more accurately address the improvement, we calculated percentage improvement which ranges from 20\% to 100\% and discussed these in Section 3.3.}

@@

\comment{(1) In the simulations, the authors should also compare the eBayes method to the method of Ji et al. (2014). Although developed for continuous microarray data, the authors could apply a simple transformation to the count data in order to make them continuous. I would be curious to know whether the eBayes method out-performs the method of Ji et al. (2014) on the transformed data.}

\response{We have assessed the approach of Ji et al. on transformed data, see Sections 3.2 and 3.3. The bottom line is that it doesn't perform as well as the methods based on our hierarchical negative-binomial model.}

@@

\comment{(2) The authors should provide more details on the computational price of the eBayes method? How long (in comparison to edgeR, baySeq, ShrinkBayes) did the eBayes method take to run? Are these methods comparable in terms of computational cost?}

\todo{We will do this.}

\comment{(3) In the description of the simulation results, the eBayes and ShrinkBayes methods display very similar performance. The distinction between the results on the Figures is quite minimal. The authors should attempt to quantify the difference in performance in a single metric. For instance, in Figure 1, take the average percent difference in TPR across the FPRs that were used to generate the plot.}

\response{The area under the curve attempts to do exactly this, i.e. a single metric, to compare the methods. We do not try to over-emphasize the difference between the ShrinkBayes and eBayes methods, which are both based on our novel model, as we do not believe there is much of a difference.}

@@

\comment{1. The authors propose a method to assess "gene expression heterosis."   While it is mentioned that this type of heterosis is a possible explanation for phenotypic heterosis, I'd like more discussion regarding this point.  The two references seem a bit outdated (for bioinformatics) so I'm wondering if this explanation is still being considered.  If it is, than I'd like to see more discussion about how the analytic results would be used to investigate this explanation.   Wouldn't scientists have specific phenotypic traits in mind and know of genes associated with these traits?  I'm a little confused why the search through the whole genome for this type of heterosis if they're looking to draw inference on the association.}

\todo{Dan? Even though some genes may be good candidates. An exploratory analysis to suggest other genes seems relevant.}

\comment{2. The model appears reasonable and is adapted from on used with microarrays.  To support their approach, they simulate data using their model (ideal case) and find that their approach does slightly better than ShrinkBayes.}

\response{We simulate data from the negative-binomial portion of the model, but do not simulate from the hierarchical portion of the model. Instead, we use \edgeR{} estimates for gene-specific parameters which provide some shrinkage for the overdispersion parameters but none for the other gene-specific parameters. We have tried to make this clear in Section 3.1. In addition, we have performed the analysis by simulating entirely from our model and achieve similar results.}

\response{We also tried to emphasize in Section 2 that ``our'' approach includes the implementation in ShrinkBayes as the model itself is novel and the eBayes and ShrinkBayes are just two different approaches to estimation.}

@@

\comment{a. How robust do you feel your method is?  There is no discussion along these lines.}

\todo{Good question. Although we suggest the gene-specific parameter distributions do not fit very well, how much better do we really think we can get if we model it well?}

\comment{b.      Why is there more of an improvement over ShrinkBayes with more replicates?  Is it because of the method used to calculate the posterior probabilities?  I'd like to see some discussion on this.}

\response{We have added discussion Section 3.3. This discussion points out the two main differences between ShrinkBayes and the eBayes approaches, but does not try to state which is the cause because, frankly, we don't know.}

### 2015-08-24 response.tex

\response{We thank the AE and two reviewers for their comments which dramatically improved the manuscript.}

Associate Editor comments:

\comment{The proposed model appears to be an extension of Ji et al. (2014) albeit with a differently specified likelihood function (negative binomial or Poisson-gamma) as perceived to be more appropriate for count data relative to the Gaussian likelihood function pursued in Ji et al. (2014). Both reviewers indicate that they are little underwhelmed by the performance of your proposed procedure relative to some of the existing methods that you compare against (i.e., ShrinkBayes, edgeR and baySeq).  Furthermore, they seem somewhat concerned about an ‘apples vs. oranges’ types of comparisons since the proposed hypothesis testing strategy in the proposed method is really not an option in the other methods.}

\response{We have assessed the Ji et al. approach and included it in Sections 3.2 and 3.3.}

@@

\comment{Page 6, top.  Is it possible then to assess your proposed method using Laplace versus normal priors to assess if that seems to be contributing to any differences between ShrinkBayes and your proposed method?  Maybe that is the real reason for the small improvements?...., and not the differences in sophistication of the constructed statistical tests?}

\response{We have assessed the use of normal distributions, see Section 3.3.}

\comment{Page 7, line 15.  Why delete the low counts?  Data quality issues?}

\todo{During the course of our research, we were initially estimating hyperparameters based on moment matching of gene-specific parameters estimated using negative binomial regression independently for all genes. This procedure had issues whenever a gene-variety combination had counts that were all zeros, thus leading to an unidentified regression parameter, or too little variability, leading to an unidentified overdispersion parameter. This cut off appeared to fix the problem. None of the current methods suffer from this problem, in particular the edgeR based parameter estimations have a built shrinkage to estimate parameters. We did not change the simulations we were using and thus report the deletion of low counts for reproducible accuracy. We do not believe removing this deletion would change any of our results.}

@@

\comment{1. The authors propose a method to assess "gene expression heterosis."   While it is mentioned that this type of heterosis is a possible explanation for phenotypic heterosis, I'd like more discussion regarding this point.  The two references seem a bit outdated (for bioinformatics) so I'm wondering if this explanation is still being considered.  If it is, than I'd like to see more discussion about how the analytic results would be used to investigate this explanation.   Wouldn't scientists have specific phenotypic traits in mind and know of genes associated with these traits?  I'm a little confused why the search through the whole genome for this type of heterosis if they're looking to draw inference on the association.}

\todo{Dan? Even though some genes may be good candidates. An exploratory analysis to suggest other genes seems relevant.}

\comment{2. The model appears reasonable and is adapted from on used with microarrays.  To support their approach, they simulate data using their model (ideal case) and find that their approach does slightly better than ShrinkBayes.}

\response{We simulate data from the negative-binomial portion of the model, but do not simulate from the hierarchical portion of the model. Instead, we use edgeR estimates for gene-specific parameters which provide some shrinkage for the overdispersion parameters but none for the other gene-specific parameters. We have tried to make this clear in Section 3.1. In addition, we have performed the analysis by simulating entirely from our model and achieve similar results.}

### 2015-08-26 ebayes_rnaseq_heterosis.tex
added computation time

Figure \ref{f:gene_specific_estimates} provides point estimates from the initial \edgeR{} estimation step and after the eBayes procedure described in Section \ref{s:ebayes}. 
\begin{figure}
\centerline{\includegraphics[width=\textwidth]{gene_specific_estimates.pdf}}
\caption{Two-dimensional histogram of point estimates from \edgeR{} and eBayes (posterior means) along with the $y=x$ diagonal.}
\label{f:gene_specific_estimates}
\end{figure}
This figure shows shrinkage for large absolute values of $\alpha$ and $\delta$ from the \edgeR{} estimation toward $\hat{\eta}_\alpha$ and $\hat{\eta}_\delta$ which are both approximately zero. The figure also shows decreased values for the overdispersion parameter with larger decreases for high and low values of overdispersion. Finally, very little, if any, shrinkage is observed for $\phi$.

@@

Although our method appears to improve upon existing methods, we believe our approach can be improved by refining the hierarchical model for the gene-specific parameter distribution. Figure \ref{f:scatterplot} shows marginal and bivariate histograms for the empirical Bayes posterior means for the gene-specific parameters. 
\begin{figure}
\centerline{\includegraphics[width=\textwidth]{estimates}}
\caption{Marginal and bivariate histograms for gene-specific parameter posterior meansfor the B73 $\times$ Mo17 maize experiment. For reference, the marginal histograms have a dashed line indicating the appropriate marginal probability density functions in the hierarchical model with the MAP estimates.}
\label{f:scatterplot}
\end{figure}
These figures show markedly departures from marginal model assumptions, e.g. normality assumptions for $\phi_g$ and $\psi_g$, and independence assumptions, e.g. $\phi_g$ versus $\psi_g$ and $\alpha_g$ versus $\delta_g$. The plot of $\phi_g$ versus $\psi_g$ shows a pattern where the mean overdispersion decreases as the mean expression level increases. The plot of $\alpha_g$ versus $\delta_g$ shows a rotated V pattern where $\delta_g$ appears to be equal to $|\alpha_g|$. This pattern is consistent with Mendel's Law of Dominance where the hybrid has mean expression equal to the parent with higher mean expression.

>>>>

We analyzed the data using all the methods compared in the previous section. The computation time on a Dell Precision T3600 with a 3.6GHz intel Xeon processor was 14 seconds for \edgeR{}, 1.3 minutes for \citeauthor{ji2014estimation}, 10 hours for the eBayes approaches, and 17 hours for \baySeq{}. In the eBayes approach, the vast majority of the time is spent on independent MCMC analysis for each gene. Thus we parallelized this step using {\tt doMC} \citep{revolution2014domc} and {\tt plyr} \citep{hadley2011split}. When parallelized across 5 cores, we the eBayes approaches took about 2.5 hours.

Figure \ref{f:gene_specific_estimates} provides point estimates from the initial \edgeR{} estimation step and after the eBayes procedure described in Section \ref{s:ebayes}. 
\begin{figure}
\centerline{\includegraphics[width=\textwidth]{gene_specific_estimates.pdf}}
\caption{Two-dimensional histogram of point estimates from \edgeR{} and posterior means from eBayes using Laplace distributions along with the $y=x$ diagonal.}
\label{f:gene_specific_estimates}
\end{figure}
This figure shows shrinkage for large absolute values of $\alpha$ and $\delta$ from the \edgeR{} estimation toward $\hat{\eta}_\alpha$ and $\hat{\eta}_\delta$ which are both approximately zero. The figure also shows decreased values for the overdispersion parameter with larger decreases for high and low values of overdispersion. Finally, very little, if any, shrinkage is observed for $\phi$.

@@

Although our method appears to improve upon existing methods, we believe our approach can be improved by refining the hierarchical model for the gene-specific parameter distribution. Figure \ref{f:scatterplot} shows marginal and bivariate histograms for the empirical Bayes posterior means for the gene-specific parameters. 
\begin{figure}
\centerline{\includegraphics[width=\textwidth]{estimates}}
\caption{Marginal and bivariate histograms of posterior means for gene-specific parameters for the B73 $\times$ Mo17 maize experiment.}
\label{f:scatterplot}
\end{figure}
These figures show markedly departures from marginal model assumptions, e.g. normality assumptions for $\phi_g$ and $\psi_g$, and independence assumptions, e.g. $\phi_g$ versus $\psi_g$ and $\alpha_g$ versus $\delta_g$. The plot of $\phi_g$ versus $\psi_g$ shows a pattern where the mean overdispersion decreases as the mean expression level increases. The plot of $\alpha_g$ versus $\delta_g$ shows a rotated V pattern where $\delta_g$ appears to be equal to $|\alpha_g|$. This pattern is consistent with Mendel's Law of Dominance where the hybrid has mean expression equal to the parent with higher mean expression.
